\section{ND-Arrays}
The command \cmdlink{narray} is a TclOO class based on the superclass \cmdlink{::ndlist::ValueContainer}.
It is an object-oriented approach to array manipulation and processing.

\begin{syntax}
\command{narray} new \$varName <\$value> <\$rank>
\end{syntax}
\begin{syntax}
narray create \$name \$varName <\$value> <\$rank>
\end{syntax}
\begin{args}
\$varName & Variable to store object name for access and garbage collection. 
Variable names are restricted to word characters and namespace delimiters only.\\
\$value & ND-list value to store in array. Default blank. \\
\$rank & Number of dimensions of array. Default ``auto''. \\
\$name & Name of object if using ``create'' method.
\end{args}
\subsection{Value Access}
The value stored in an narray object can be accessed by calling the object by itself with no argument, or by calling the method ``value''.
\begin{syntax}
\method{narray}{<value>}
\end{syntax}
\begin{args}
value & Optional method for returning value.
\end{args}
\begin{example}{Creating an ``narray'' object and accessing the value}
\begin{lstlisting}
narray new x {hello world}
puts [$x]
\end{lstlisting}
\tcblower
\begin{lstlisting}
hello world
\end{lstlisting}
\end{example}

\clearpage
\subsection{Number of Dimensions}
The number of dimensions can be manually set when creating the narray object, but by default it is set to ``auto''. For numerical data that is properly formatted as an ND-list, it will always choose the minimum number of dimensions that can store the data. For data containing strings with spaces, it is advised to manually set the dimension. To query or change the number of dimensions, use the method \methodlink[0]{narray}{ndims}.
\begin{syntax}
\method{narray}{ndims} <\$rank | auto | type>
\end{syntax}
\begin{args}
\$rank & Value to set as number of dimensions of array. Default blank to query. \\
auto & Toggle auto-dimensions. Returns auto-determined number of dimensions. \\
type & Query internal dimension type. Returns ``auto'' or ``user''.
\end{args}
\subsection{Shape and Size}
The shape and size corresponding with the number of dimensions of the narray are accessed with the methods \methodlink[0]{narray}{shape} and \methodlink[0]{narray}{size}.
\begin{syntax}
\method{narray}{shape}
\end{syntax}
\begin{syntax}
\method{narray}{size}
\end{syntax}
\begin{example}{N-array dimension, shape and size}
\begin{lstlisting}
narray new x {{1 2 3} {4 5 6}}
puts [list [$x ndims] [$x ndims type] [$x shape] [$x size]]
$x ndims 3; # change number of dimensions
puts [list [$x ndims] [$x ndims type] [$x shape] [$x size]]
\end{lstlisting}
\tcblower
\begin{lstlisting}
2 auto {2 3} 6
3 user {2 3 1} 6
\end{lstlisting}
\end{example}

\clearpage
\subsection{Indexing}
The ``\texttt{@}'' operator uses \cmdlink{nget} to access a portion of the ND-array. 
\begin{syntax}
\index{narray methods!@} \$narrayObj @ \$i ...
\end{syntax}
\begin{args}
\$i ... & Index inputs corresponding with rank of ND-array. \\
\end{args}

\begin{example}{Accessing portions of an ND-array}
\begin{lstlisting}
narray new x {{1 2 3} {4 5 6} {7 8 9}}
puts [$x @ 0 2]
puts [$x @ 0:end-1 {0 2}]
\end{lstlisting}
\tcblower
\begin{lstlisting}
3
{1 3} {4 6}
\end{lstlisting}
\end{example}

\subsection{Copying}
The operator ``\texttt{-{}->}'' copies the ND-array to a new variable, and returns the new object.
If indices are specified, the new ND-array object will have the rank of the indexed range.
\begin{syntax}
\index{narray methods!-{}->} \$narrayObj <@ \$i ...> -{}-> \$varName
\end{syntax}
\begin{args}
\$i ... & Indices to access. Default all. \\
\$varName & Variable to store object name for access and garbage collection. 
Variable names are restricted to word characters and namespace delimiters only.
\end{args}

\begin{example}{Copying a portion of an ND-array}
\begin{lstlisting}
narray new x {{1 2 3} {4 5 6}}
$x @ 0* : --> y; # Row vector (flattened to 1D)
puts "[$y ndims], [$y]"
\end{lstlisting}
\tcblower
\begin{lstlisting}
1, 1 2 3
\end{lstlisting}
\end{example}
\clearpage
\subsection{Evaluation/Mapping}
The command \cmdlink{neval} maps over ND-arrays using \cmdlink{nmap}. 
The command \cmdlink{nexpr} is a special case that passes input through the Tcl \textit{expr} command.
ND-arrays can be referred to with ``\texttt{@ref}'', where ``ref'' is the name of the ND-array variable.
Portions of an ND-array can be mapped over with the notation ``\texttt{@ref(\$i,...)}''.
The highest-rank input will determine the output, and all input arrays must have compatible dimensions.
\begin{syntax}
\command{neval} \$body <\$self> <\$rankVar>
\end{syntax}
\begin{syntax}
\command{nexpr} \$expr <\$self> <\$rankVar>
\end{syntax}
\begin{args}
\$body & Script to evaluate, with ``\texttt{@ref}'' notation for object references. \\
\$expr & Expression to evaluate, with ``\texttt{@ref}'' notation for object references. \\
\$self & Object to refer to with ``\texttt{@.}''. Default blank. \\
\$rankVar & Variable to store resulting rank in. Default blank.
\end{args}

\begin{example}{Get distance between elements in a vector}
\begin{lstlisting}
narray new x {1 2 4 7 11 16}
puts [nexpr {@x(1:end) - @x(0:end-1)}]
\end{lstlisting}
\tcblower
\begin{lstlisting}
1 2 3 4 5
\end{lstlisting}
\end{example}

\begin{example}{Outer product of two vectors}
\begin{lstlisting}
narray new x {1 2 3}
narray new y {{4 5 6}}
puts [nexpr {@x * @y}]
\end{lstlisting}
\tcblower
\begin{lstlisting}
{4 5 6} {8 10 12} {12 15 18}
\end{lstlisting}
\end{example}

\clearpage
\subsection{Modification}
The assignment operator, ``\texttt{=}'', sets the value of the entire ND-array, or of a specified range.
The math assignment operator, ``\texttt{:=}'', sets the value, passing the input through the \cmdlink{nexpr} command. 
Both assignment operators return the object.

\begin{syntax}
\index{narray methods!=} \$narrayObj <@ \$i ...> = \$value
\end{syntax}
\begin{syntax}
\index{narray methods!:=} \$narrayObj <@ \$i ...> := \$expr
\end{syntax}
\begin{args}
\$i ... & Indices to modify. Default all. \\
\$value & Value to assign. Blank to remove values. \\
\$expr & Expression to evaluate.
\end{args}

If using the math assignment operator, the ND-array or indexed range can be accessed with the alias ``\texttt{\$.}'', and the elements of the array or indexed range can be accessed with ``\texttt{@.}''.
\begin{syntax}
\$. \$arg ...
\end{syntax}
\begin{args}
\$arg ... & Method arguments for object.
\end{args}

\begin{example}{Element-wise modification of a vector}
\begin{lstlisting}
# Create blank vectors and assign values
[narray new x] = {1 2 3}
[narray new y] = {10 20 30}
# Add one to each element
puts [[$x := {@. + 1}]]
# Double the last element
puts [[$x @ end := {@. * 2}]]
# Element-wise addition of vectors
puts [[$x := {@. + @y}]]
\end{lstlisting}
\tcblower
\begin{lstlisting}
2 3 4
2 3 8
12 23 38
\end{lstlisting}
\end{example}

\clearpage
\subsection{Removal/Insertion}
The method \methodlink[0]{narray}{remove} removes portions of an ND-array along a specified axis with the command \cmdlink{nremove}, and the method \methodlink[0]{narray}{insert} inserts values into an ND-array at a specified index/axis with the command \cmdlink{ninsert}. 
Both methods modify the object and return the object. 

\begin{syntax}
\method{narray}{remove} \$i <\$axis>
\end{syntax}
\begin{syntax}
\method{narray}{insert} \$i \$sublist <\$axis>
\end{syntax}
\begin{args}
\$i & Indices to remove/insert at. \\
\$sublist & Value to insert. \\
\$axis & Axis to remove/insert at (default 0).
\end{args}

\begin{example}{Removing elements from a vector}
\begin{lstlisting}
narray new vector {1 2 3 4 5 6 7 8}
# Remove all odd numbers
$vector remove [find [nexpr {@vector % 2}]]
puts [$vector]
\end{lstlisting}
\tcblower
\begin{lstlisting}
2 4 6 8
\end{lstlisting}
\end{example}

\begin{example}{Inserting a column into a matrix}
\begin{lstlisting}
narray new matrix {{1 2} {3 4} {5 6}}
$matrix insert 1 {A B C} 1
puts [$matrix]
\end{lstlisting}
\tcblower
\begin{lstlisting}
{1 A 2} {3 B 4} {5 C 6}
\end{lstlisting}
\end{example}

\clearpage
\subsection{Temporary Object Evaluation}
The pipe operator, ``\texttt{|}'', copies the ND-array to a temporary object, and evaluates the method.
Returns the result of the method, or the value of the temporary object.
This operator is useful for converting methods that modify the object to methods that return a modified value.
\begin{syntax}
\index{narray methods!$\vert$} \$narrayObj <@ \$i ...> | \$method \$arg ...
\end{syntax}
\begin{args}
\$i ... & Indices to access. Default all. \\
\$method & Method to evaluate. \\
\$arg ... & Arguments to pass to method.
\end{args}
\begin{example}{Temporary object value}
\begin{lstlisting}
# Create a matrix
narray new x {{1 2 3} {4 5 6}}
# Print value with first row doubled.
puts [$x | @ 0* : := {@. * 2}]
# Source object was not modified
puts [$x]
\end{lstlisting}
\tcblower
\begin{lstlisting}
{2 4 6} {4 5 6}
{1 2 3} {4 5 6}
\end{lstlisting}
\end{example}
\clearpage
\subsection{Reference Variable Evaluation}
The ampersand operator ``\texttt{\&}'' copies the ND-array value or range to a reference variable, and evaluates a body of script. 
The changes made to the reference variable will be applied to the object, and if the variable is unset, the object will be deleted.
If no indices are specified and the variable is unset in the script, the ND-array object will be destroyed.
Returns the result of the script.

\begin{syntax}
\index{narray methods!\&} \$narrayObj <@ \$i ...> \& \$refName \$body
\end{syntax}
\begin{args}
\$i ... & Indices to access. Default all. \\
\$refName & Variable name to use for reference. \\
\$body & Body to evaluate.
\end{args}

\begin{example}{Appending a vector}
\begin{lstlisting}
# Create a 1D list
narray new x {1 2 3}
# Append the list
$x & ref {lappend ref 4 5 6}
puts [$x]
# Append a subset of the list
$x @ end* & ref {lappend ref 7 8 9}
puts [$x]
\end{lstlisting}
\tcblower
\begin{lstlisting}
1 2 3 4 5 6
1 2 3 4 5 {6 7 8 9}
\end{lstlisting}
\end{example}

\clearpage


\clearpage


