\section{N-Dimensional Lists (Tensors)}
All Tcl values are ND-lists. An ND-list is defined as a list of equal length (N-1)D-lists, which are defined as equal length (N-2)D-lists, and so on until 0D-lists, which are simply strings that either have no list representation or are of list length 1.
This definition is flexible, and allows for different interpretations of the same data. 
For example, the list ``1 2 3'' can be interpreted as a scalar with value ``1 2 3'', a vector with values ``1'', ``2'', and ``3'', or a matrix with row vectors ``1'', ``2'', and ``3''. 
In general, if a value is a valid for N dimensions, it will also be valid for dimensions 0 to N-1.

The command \cmdlink{ndims} returns the rank of an ND-list, and the command \cmdlink{ndims\_multiple} returns the rank that is compatible with multiple ND-lists. 
By default, it automatically determines the rank for the data, but if a rank is provided, it will validate that the ND-list or ND-lists are compatible with the provided rank. 
\begin{syntax}
\command{ndims} \$ndlist <\$rank>
\end{syntax}
\begin{syntax}
\command{ndims\_multiple} \$ndlists <\$rank>
\end{syntax}
\begin{args}
\$ndlist & ND-list. \\
\$ndlists & List of ND-lists. \\
\$rank & Rank of ND-list (e.g. 2 for matrix) or ``auto'' for auto-rank. Default ``auto''.
\end{args}

\begin{example}{Rank of an ND-list}
\begin{lstlisting}
set x {1}
set y {1 2 {hello world}}; # note that this is not a valid 2D list
set z {{1 2 3} {4 5 6}}
puts [ndims $x]; # 0
puts [ndims $y]; # 1
puts [ndims $z]; # 2
# the only rank that works for x, y, and z is 1
puts [ndims_multiple [list $x $y $z]]; # 1
\end{lstlisting}
\tcblower
\begin{lstlisting}
0
1
2
1
\end{lstlisting}
\end{example}

\clearpage
\subsection{Shape and Size}
The commands \cmdlink{nshape} and \cmdlink{nsize} return the shape and size of an ND-list, respectively.
The shape is a list of the dimensions, and the size is the product of the shape.
\begin{syntax}
\command{nshape} \$ndlist <\$rank>
\end{syntax}
\begin{syntax}
\command{nsize} \$ndlist <\$rank>
\end{syntax}
\begin{args}
\$ndlist & ND-list to get shape/size of. \\
\$rank & Rank of ND-list (e.g. 2 for matrix) or ``auto'' for auto-rank. Default ``auto''.
\end{args}
\begin{example}{Getting shape and size of an ND-list}
\begin{lstlisting}
# Create a 3D list
set x {{{1 2} {3 4} {5 6}} {{7 8} {9 10} {11 12}}}
# Get the shape and size for different rank interpretations
puts [list [nshape $x] [nsize $x]]; # auto-rank (3)
puts [list [nshape $x 1] [nsize $x 1]]; # rank 1
puts [list [nshape $x 2] [nsize $x 2]]; # rank 2
puts [list [nshape $x 3] [nsize $x 3]]; # rank 3
puts [list [nshape $x 4] [nsize $x 4]]; # rank 4
\end{lstlisting}
\tcblower
\begin{lstlisting}
{2 3 2} 12
2 2
{2 3} 6
{2 3 2} 12
{2 3 2 1} 12
\end{lstlisting}
\end{example}

\clearpage
\subsection{Initialization}
The command \cmdlink{nfull} initializes a valid ND-list of any size filled with a single value.
\begin{syntax}
\command{nfull} \$value \$shape
\end{syntax}
\begin{args}
\$value & Value to repeat \\
\$shape & Shape (list of dimensions) of ND-list. 
\end{args}
\begin{example}{Generate ND-list filled with one value}
\begin{lstlisting}
puts [nfull foo {3 2}]; # 3x2 matrix filled with "foo"
puts [nfull 0 {2 2 2}]; # 2x2x2 tensor filled with zeros
\end{lstlisting}
\tcblower
\begin{lstlisting}
{foo foo} {foo foo} {foo foo}
{{0 0} {0 0}} {{0 0} {0 0}}
\end{lstlisting}
\end{example}
The command \cmdlink{nrand} initializes a valid ND-list of any size filled with random values between 0 and 1.
\begin{syntax}
\command{nrand} \$shape
\end{syntax}
\begin{args}
\$shape & Shape (list of dimensions) of ND-list. 
\end{args}
\begin{example}{Generate random matrix}
\begin{lstlisting}
expr {srand(0)}; # resets the random number seed (for the example)
puts [nrand {1 2}]; # 1x2 matrix filled with random numbers
\end{lstlisting}
\tcblower
\begin{lstlisting}
{0.013469574513598146 0.3831388500440581}
\end{lstlisting}
\end{example}
\clearpage
\subsection{Repeating and Expanding}
The command \cmdlink{nrepeat} repeats portions of an ND-list a specified number of times.
\begin{syntax}
\command{nrepeat} \$ndlist \$repeats
\end{syntax}
\begin{args}
\$value & Value to repeat \\
\$repeats & Repetitions at each level.
\end{args}
\begin{example}{Repeat elements of a matrix}
\begin{lstlisting}
puts [nrepeat {{1 2} {3 4}} {1 2}]
\end{lstlisting}
\tcblower
\begin{lstlisting}
{1 2 1 2} {3 4 3 4}
\end{lstlisting}
\end{example}
The command \cmdlink{nexpand} repeats portions of an ND-list to expand to new dimensions.
New dimensions must be divisible by old dimensions.
For example, 1x1, 2x1, 4x1, 1x3, 2x3 and 4x3 are compatible with 4x3.
\begin{syntax}
\command{nexpand} \$ndlist \$shape
\end{syntax}
\begin{args}
\$ndlist & ND-list to expand. \\
\$shape & New shape of ND-list. If -1 is used, it keeps that axis the same.
\end{args}
\begin{example}{Expand an ND-list to new dimensions}
\begin{lstlisting}
puts [nexpand {1 2 3} {-1 2}]
puts [nexpand {{1 2}} {2 4}]
\end{lstlisting}
\tcblower
\begin{lstlisting}
{1 1} {2 2} {3 3}
{1 2 1 2} {1 2 1 2}
\end{lstlisting}
\end{example}
\clearpage
\subsection{Padding and Extending}
The command \cmdlink{npad} pads an ND-list along its axes by a specified number of elements.
\begin{syntax}
\command{npad} \$ndlist \$value \$pads
\end{syntax}
\begin{args}
\$ndlist & ND-list to pad. \\
\$value & Value to pad with. \\
\$pads & Number of elements to pad along each axis. Negative to prepend.
\end{args}
\begin{example}{Padding an ND-list with a value}
\begin{lstlisting}
set a {{1 2 3} {4 5 6} {7 8 9}}
puts [npad $a 0 {2 1}]
puts [npad $a foo {-2 -1}]
\end{lstlisting}
\tcblower
\begin{lstlisting}
{1 2 3 0} {4 5 6 0} {7 8 9 0} {0 0 0 0} {0 0 0 0}
{foo foo foo foo} {foo foo foo foo} {foo 1 2 3} {foo 4 5 6} {foo 7 8 9}
\end{lstlisting}
\end{example}
The command \cmdlink{nextend} extends an ND-list to a new shape by padding.
\begin{syntax}
\command{nextend} \$ndlist \$value \$shape
\end{syntax}
\begin{args}
\$ndlist & ND-list to extend. \\
\$value & Value to pad with. \\
\$shape & New shape of ND-list. To keep the shape at an axis, use -1.
\end{args}
\begin{example}{Extending an ND-list to a new shape with a filler value}
\begin{lstlisting}
set a {hello hi hey howdy}
puts [nextend $a world {-1 2}]; # -1 preserves size at axis 0
\end{lstlisting}
\tcblower
\begin{lstlisting}
{hello world} {hi world} {hey world} {howdy world}
\end{lstlisting}
\end{example}
\clearpage
\subsection{Flattening and Reshaping}
The command \cmdlink{nflatten} flattens an ND-list to a vector.
\begin{syntax}
\command{nflatten} \$ndlist <\$rank>
\end{syntax}
\begin{args}
\$ndlist & ND-list to flatten. \\
\$rank & Rank of ND-list (e.g. 2 for matrix) or ``auto'' for auto-rank. Default ``auto''.
\end{args}
\begin{example}{Flattening a 3D tensor}
\begin{lstlisting}
puts [nflatten {{{1 2} {3 4}} {{5 6} {7 8}}}]
\end{lstlisting}
\tcblower
\begin{lstlisting}
1 2 3 4 5 6 7 8
\end{lstlisting}
\end{example}

The command \cmdlink{nreshape} reshapes a vector into specified dimensions.
Sizes must be compatible.
\begin{syntax}
\command{nreshape} \$vector \$shape
\end{syntax}
\begin{args}
\$vector & Vector (1D-list) to reshape. \\
\$shape & Shape of ND-list. One axis may be dynamic, denoted with -1.
\end{args}
\begin{example}{Reshape a vector to a matrix with three columns}
\begin{lstlisting}
puts [nreshape {1 2 3 4 5 6} {-1 3}]
\end{lstlisting}
\tcblower
\begin{lstlisting}
{1 2 3} {4 5 6}
\end{lstlisting}
\end{example}


\clearpage

\subsection{Index Notation}
This package provides generalized N-dimensional list access/modification commands, using an index notation parsed by the command \cmdlink{::ndlist::ParseIndex}, which returns the index type and an index list for the type.
\begin{syntax}
\command{::ndlist::ParseIndex} \$n \$input
\end{syntax}
\begin{args}
\$n & Number of elements in list. \\
\$input & Index input. Options are shown below: \\
\quad : & All indices \\
\quad \$start:\$stop & Range of indices (e.g. 0:4 or 1:end-2).\\
\quad \$start:\$step:\$stop & Stepped range of indices (e.g. 0:2:-2 or 2:3:end). \\
\quad \$iList & List of indices (e.g. \{0 end-1 5\} or 3). \\
\quad \$i* & Single index with a asterisk, ``flattens'' the ndlist (e.g. 0* or end-3*). 
\end{args}
Additionally, indices get passed through the \cmdlink{::ndlist::Index2Integer} command, which converts the inputs ``end'', ``end-integer'', ``integer$\pm$integer'' and negative wrap-around indexing (where -1 is equivalent to ``end'') into normal integer indices.
Note that this command will return an error if the index is out of range.
\begin{syntax}
\command{::ndlist::Index2Integer} \$n \$index
\end{syntax}
\begin{args}
\$n & Number of elements in list. \\
\$index & Single index. 
\end{args}

\begin{example}{Index Notation}
\begin{lstlisting}
set n 10
puts [::ndlist::ParseIndex $n :]
puts [::ndlist::ParseIndex $n 1:8]
puts [::ndlist::ParseIndex $n 0:2:6]
puts [::ndlist::ParseIndex $n {0 5 end-1}]
puts [::ndlist::ParseIndex $n end*]
\end{lstlisting}
\tcblower
\begin{lstlisting}
A {}
R {1 8}
L {0 2 4 6}
L {0 5 8}
S 9
\end{lstlisting}
\end{example}
\clearpage
\subsection{Access}
Portions of an ND-list can be accessed with the command \cmdlink{nget}, using the index parser \cmdlink{::ndlist::ParseIndex} for each dimension being indexed.
Note that unlike the Tcl \textit{lindex} and \textit{lrange} commands, \cmdlink{nget} will return an error if the indices are out of range.
\begin{syntax}
\command{nget} \$ndlist \$i ...
\end{syntax}
\begin{args}
\$ndlist & ND-list value. \\
\$i ... & Index inputs, parsed with \cmdlink{::ndlist::ParseIndex}. 
\end{args}
\begin{example}{ND-list access}
\begin{lstlisting}
set A {{1 2 3} {4 5 6} {7 8 9}}
puts [nget $A 0 :]; # get row matrix
puts [nget $A 0* :]; # flatten row matrix to a vector
puts [nget $A 0:1 0:1]; # get matrix subset
puts [nget $A end:0 end:0]; # can have reverse ranges
puts [nget $A {0 0 0} 1*]; # can repeat indices
# TIP: You can use the Tcl 'interp alias' command to create a convenient shortcut to nget
interp alias {} @ {} nget
puts [@ $A 0 :]; # get row matrix
\end{lstlisting}
\tcblower
\begin{lstlisting}
{1 2 3}
1 2 3
{1 2} {4 5}
{9 8 7} {6 5 4} {3 2 1}
2 2 2
{1 2 3}
\end{lstlisting}
\end{example}

\clearpage
\subsection{Modification}
A ND-list can be modified by reference with \cmdlink{nset}, and by value with \cmdlink{nreplace}, using the index parser \cmdlink{::ndlist::ParseIndex} for each dimension being indexed.
Note that unlike the Tcl \textit{lset} and \textit{lreplace} commands, the commands \cmdlink{nset} and \cmdlink{nreplace} will return an error if the indices are out of range.
If all the index inputs are ``\texttt{:}'' except for one, and the replacement list is blank, it will delete values along that axis by calling \cmdlink{nremove}.
Otherwise, the replacement ND-list must be expandable to the target index dimensions. 

If using the ``\texttt{= \$expr}'' notation, it will call \cmdlink{nexpr}, where ``self'' is the ND-list or indexed range being modified and ``rank'' is ``auto'' if no index arguments are provided, otherwise it is equal to the number of index arguments minus the number of single index (S) arguments (see \cmdlink{::ndlist::ParseIndex}).
\begin{syntax}
\command{nset} \$varName \$i ... (\$sublist | = \$expr)
\end{syntax}
\begin{syntax}
\command{nreplace} \$ndlist \$i ... (\$sublist | = \$expr)
\end{syntax}
\begin{args}
\$varName & Variable that contains an ND-list. \\
\$ndlist & ND-list to modify. \\
\$i ... & Index inputs, parsed with \cmdlink{::ndlist::ParseIndex}. \\
\$sublist & Replacement list, or blank to delete values. \\
\$expr & Expression to evaluate and replace values with. Indexed range can be accessed with ``\texttt{@.}'' for convenience.
\end{args}


\begin{example}{ND-list modification}
\begin{lstlisting}
# Swap rows in a matrix (by reference)
set a {{1 2 3} {4 5 6} {7 8 9}}
nset a {1 0} : [nget $a {0 1} :]
puts $a
# Element-wise operation on portion of vector (by value)
set b [range 10]
puts [nreplace $b 0:2:end = {@. + 10}]
\end{lstlisting}
\tcblower
\begin{lstlisting}
{4 5 6} {1 2 3} {7 8 9}
10 1 12 3 14 5 16 7 18 9
\end{lstlisting}
\end{example}
\clearpage

\clearpage
\subsection{Removal}
The command \cmdlink{nremove} removes portions of an ND-list at a specified axis.
\begin{syntax}
\command{nremove} \$ndlist \$i <\$axis>
\end{syntax}
\begin{args}
\$ndlist & ND-list to modify. \\
\$i & Index input, parsed with \cmdlink{::ndlist::ParseIndex}. \\
\$axis & Axis to remove at. Default 0.
\end{args}

\begin{example}{Filtering a list by removing elements}
\begin{lstlisting}
set x [range 10]
puts [nremove $x [where $x > 4]]
\end{lstlisting}
\tcblower
\begin{lstlisting}
0 1 2 3 4
\end{lstlisting}
\end{example}

\begin{example}{Deleting a column from a matrix}
\begin{lstlisting}
set a {{1 2 3} {4 5 6} {7 8 9}}
puts [nremove $a 2 1]
\end{lstlisting}
\tcblower
\begin{lstlisting}
{1 2} {4 5} {7 8}
\end{lstlisting}
\end{example}
\clearpage
\subsection{Insertion and Concatenation}
The command \cmdlink{ninsert} inserts an ND-list into another ND-list at a specified index and axis.
The ND-lists must agree in dimension at all other axes.
If ``end'' or ``end-integer'' is used for the index, it will insert after the index. 
Otherwise, it will insert before the index.
The command \cmdlink{ncat} is shorthand for inserting at ``end'', and concatenates two ND-lists.
\begin{syntax}
\command{ninsert} \$ndlist1 \$index \$ndlist2 <\$axis> <\$rank>
\end{syntax}
\begin{syntax}
\command{ncat} \$ndlist1 \$ndlist2 <\$axis> <\$rank> 
\end{syntax}
\begin{args}
\$ndlist1 \$ndlist2 & ND-lists to combine. \\
\$index & Index to insert at. \\
\$axis & Axis to insert/concatenate at (default 0). \\
\$rank & Rank of ND-list (e.g. 2 for matrix) or ``auto'' for auto-rank. Default ``auto''.
\end{args}

\begin{example}{Inserting a column into a matrix}
\begin{lstlisting}
set matrix {{1 2} {3 4} {5 6}}
set column {A B C}
puts [ninsert $matrix 1 $column 1 2]
\end{lstlisting}
\tcblower
\begin{lstlisting}
{1 A 2} {3 B 4} {5 C 6}
\end{lstlisting}
\end{example}
\begin{example}{Concatenate tensors}
\begin{lstlisting}
set x [nreshape {1 2 3 4 5 6 7 8 9} {3 3 1}]
set y [nreshape {A B C D E F G H I} {3 3 1}]
puts [ncat $x $y 2 3]
\end{lstlisting}
\tcblower
\begin{lstlisting}
{{1 A} {2 B} {3 C}} {{4 D} {5 E} {6 F}} {{7 G} {8 H} {9 I}}
\end{lstlisting}
\end{example}

\clearpage
\subsection{Changing Order of Axes}
The command \cmdlink{nswapaxes} is a general purpose transposing function that swaps the axes of an ND-list.
For simple matrix transposing, the command \cmdlink{transpose} can be used instead.
\begin{syntax}
\command{nswapaxes} \$ndlist \$axis1 \$axis2
\end{syntax}
\begin{args}
\$ndlist & ND-list to manipulate. \\
\$axis1 \$axis2 & Axes to swap.
\end{args}
The command \cmdlink{nmoveaxis} moves a specified source axis to a target position. 
For example, moving axis 0 to position 2 would change ``i,j,k'' to ``j,k,i''.
\begin{syntax}
\command{nmoveaxis} \$ndlist \$source \$target
\end{syntax}
\begin{args}
\$ndlist & ND-list to manipulate. \\
\$source & Source axis. \\
\$target & Target position.
\end{args}
The command \cmdlink{npermute} is more general purpose, and defines a new order for the axes of an ND-list.
For example, the axis list ``1 0 2'' would change ``i,j,k'' to ``j,i,k''. 
\begin{syntax}
\command{npermute} \$ndlist \$order
\end{syntax}
\begin{args}
\$ndlist & ND-list to manipulate. \\
\$order & List of axes defining new order.
\end{args}
\begin{example}{Changing tensor axes}
\begin{lstlisting}
set x {{{1 2} {3 4}} {{5 6} {7 8}}}; # 3D tensor
set y [nswapaxes $x 0 2]; # k,j,i
set z [nmoveaxis $x 0 2]; # j,k,i
puts [lindex $x 0 0 1]
puts [lindex $y 1 0 0]
puts [lindex $z 0 1 0]
\end{lstlisting}
\tcblower
\begin{lstlisting}
2
2
2
\end{lstlisting}
\end{example}

\clearpage
\subsection{ND Functional Mapping}
The command \cmdlink{napply} applies a command over each element of an ND-list, and returns the result.
The commands \cmdlink{napply2} maps element-wise over two ND-lists. 
If the input lists have different shapes, they will be expanded to their maximum dimensions with \cmdlink{nexpand} (if compatible).
\begin{syntax}
\command{napply} \$command \$ndlist <\$suffix> <\$rank>
\end{syntax}
\begin{syntax}
\command{napply2} \$command \$ndlist1 \$ndlist2 <\$suffix> <\$rank> 
\end{syntax}
\begin{args}
\$ndlist & ND-list to map over. \\
\$ndlist1 \$ndlist2 & ND-lists to map over, element-wise. \\
\$command & Command prefix to map with. \\
\$suffix & Additional arguments to append after ND-list elements. Default blank. \\
\$rank & Rank of ND-list (e.g. 2 for matrix) or ``auto'' for auto-rank. Default ``auto''.
\end{args}
\begin{example}{Chained functional mapping over a matrix}
\begin{lstlisting}
napply puts [napply {format %.2f} [napply expr {{1 2} {3 4}} {+ 1}]]
\end{lstlisting}
\tcblower
\begin{lstlisting}
2.00
3.00
4.00
5.00
\end{lstlisting}
\end{example}
\begin{example}{Format columns of a matrix}
\begin{lstlisting}
set data {{1 2 3} {4 5 6} {7 8 9}}
set formats {{%.1f %.2f %.3f}}
puts [napply2 format $formats $data]
\end{lstlisting}
\tcblower
\begin{lstlisting}
{1.0 2.00 3.000} {4.0 5.00 6.000} {7.0 8.00 9.000}
\end{lstlisting}
\end{example}
\clearpage
\subsection{Reducing an ND-list}
The command \cmdlink{nreduce} combines \cmdlink{nmoveaxis} and \cmdlink{napply} to reduce an axis of an ND-list with a function that reduces a vector to a scalar, like \cmdlink{max} or \cmdlink{sum}.
\begin{syntax}
\command{nreduce} \$command \$ndlist <\$axis> <\$suffix> <\$rank>
\end{syntax}
\begin{args}
\$command & Command prefix to map with. \\
\$ndlist & ND-list to map over. \\
\$axis & Axis to reduce. Default 0. \\
\$suffix & Additional arguments to append after ND-list elements. Default blank. \\
\$rank & Rank of ND-list (e.g. 2 for matrix) or ``auto'' for auto-rank. Default ``auto''.
\end{args}
\begin{example}{Matrix row and column statistics}
\begin{lstlisting}
set x {{1 2} {3 4} {5 6} {7 8}}
puts [nreduce max $x]; # max of each column
puts [nreduce max $x 1]; # max of each row
puts [nreduce sum $x]; # sum of each column
puts [nreduce sum $x 1]; # sum of each row
\end{lstlisting}
\tcblower
\begin{lstlisting}
7 8
2 4 6 8
16 20
3 7 11 15
\end{lstlisting}
\end{example}

\clearpage
\subsection{Generalized N-Dimensional Mapping}
The command \cmdlink{nmap} is a general purpose mapping function for N-dimensional lists in Tcl.
If multiple ND-lists are provided for iteration, they must be expandable to their maximum dimensions.
The actual implementation flattens all the ND-lists and calls the Tcl \textit{lmap} command, and then reshapes the result to the target dimensions.
So, if ``continue'' or ``break'' are used in the map body, it will return an error.
If rank is not specified, it will automatically determine the rank using \cmdlink{ndims\_multiple} and the referenced ND-lists.
\begin{syntax}
\command{nmap} <\$rank> \$varName \$ndlist <\$varName \$ndlist ...> \$body
\end{syntax}
\begin{args}
\$rank & Rank (e.g. 2 for matrix) or ``auto'' for auto-rank. Default ``auto''.  \\
\$varName & Variable name to iterate with. \\
\$ndlist & ND-list to iterate over. \\
\$body & Tcl script to evaluate at every loop iteration. 
\end{args}

\begin{example}{Expand and map over matrices}
\begin{lstlisting}
set phrases [nmap 2 greeting {{hello goodbye}} subject {world moon} {
    list $greeting $subject
}]
napply puts $phrases {} 2 
\end{lstlisting}
\tcblower
\begin{lstlisting}
hello world
goodbye world
hello moon
goodbye moon
\end{lstlisting}
\end{example}
\clearpage
\subsection{Loop Index Access}
The iteration indices of \cmdlink{nmap} can be accessed with the commands \cmdlink{i}, \cmdlink{j}, and \cmdlink{k}. 
The commands \cmdlink{j} and \cmdlink{k} are simply shorthand for \cmdlink{i} with axes 1 and 2.
\begin{syntax}
\command{i} <\$axis>
\end{syntax}
\begin{syntax}
\command{j}
\end{syntax}
\begin{syntax}
\command{k}
\end{syntax}
\begin{args}	
\$axis & Dimension to access mapping index at. Default 0. \\
 & If -1, returns the linear index of the loop.
\end{args}

\begin{example}{Finding index tuples that match criteria}
\begin{lstlisting}
set x {{1 2 3} {4 5 6} {7 8 9}}
set indices {}
nmap xi $x {
    if {$xi > 4} {
        lappend indices [list [i] [j]]
    }
}
puts $indices
\end{lstlisting}
\tcblower
\begin{lstlisting}
{1 1} {1 2} {2 0} {2 1} {2 2}
\end{lstlisting}
\end{example}
\clearpage

\subsection{Element-Wise Expressions}
The command \cmdlink{neval} maps over ND-arrays using \cmdlink{nmap}, but without the need to specify looping variables, and the command \cmdlink{nexpr} is a special case that passes input through the Tcl \textit{expr} command.
ND-lists can be referred to with ``\texttt{@ref}'', where ``ref'' is the name of the variable storing the ND-list.
If an ND-list is provided as ``self'', it can be referred to with ``\texttt{@.}'' for convenience.
Additionally, portions of an ND-list can be mapped over with the notation ``\texttt{@ref(\$i,...)}'', where ``\texttt{\$i,...}'' are raw index arguments (does not do any substitution).
If rank is not specified, it will automatically determine the rank using \cmdlink{ndims\_multiple} and the referenced/indexed ND-lists.
\begin{syntax}
\command{neval} \$body <\$self> <\$rank>
\end{syntax}
\begin{syntax}
\command{nexpr} \$expr <\$self> <\$rank>
\end{syntax}
\begin{args}
\$body & Script to evaluate, with ``\texttt{@ref}'' notation for ND-list references. \\
\$expr & Expression to evaluate, with ``\texttt{@ref}'' notation for ND-list references. \\
\$self & ND-list to refer to with ``\texttt{@.}''. Default blank. \\
\$rank & Rank (e.g. 2 for matrix) or ``auto'' for auto-rank. Default ``auto''. 
\end{args}

\begin{example}{Element-wise operations}
\begin{lstlisting}
set x {1 2 3}; # vector, length 3
set y {{4 5 6}}; # matrix, shape {1 3}
set z {1 2 4 7 11 16}; # vector, length 6
puts [nexpr {@x * @y}]; # outer product of two vectors (creates matrix)
puts [nexpr {@x + @z}]; # expands vector x to match length of z
puts [nexpr {@z(1:end) - @z(0:end-1)}]; # distance between vector elements
puts [nexpr {@. * 2.0} {4 3 8}]; # self-operation using @. notation
\end{lstlisting}
\tcblower
\begin{lstlisting}
{4 5 6} {8 10 12} {12 15 18}
2 4 7 8 13 19
1 2 3 4 5
8.0 6.0 16.0
\end{lstlisting}
\end{example}

\clearpage
\subsection{Prefix Notation Math}
Tcl is unique in that everything is a command. So, that means that it does not natively support the typical algebraic ``infix'' notation that people are used to. 
Instead, the Tcl \textit{expr} command provides a sub-language that parses scalar math in the familiar ``infix'' notation.
The ndlist \cmdlink{nexpr} command was designe to extend the functionality of \textit{expr} to provide support for element-wise operations. 
However, in code that is heavy in math, having a lot of ``expr'' or ``nexpr'' calls can make code difficult to read.
There has been a lot of discussion in the Tcl community about this problem, and the general consensus is that it was addressed in Tcl \hyperlink{https://core.tcl-lang.org/tips/doc/trunk/tip/174.md}{TIP \#174}, which added the ::tcl::mathop namespace to the core. 
This change to Tcl provided users with easy access to math operators as commands.
The ndlist package expands upon this by providing equivalent element-wise operators as commands, with the exception that it only provides element-wise math operators. 
This is because the ndlist math operator commands do not have an option to specify the rank of the input lists, which could lead to unexpected results for some string/list operations.

The available element-wise operators are \texttt{$\sim$}, \texttt{!}, \texttt{-}, \texttt{+}, \texttt{*}, \texttt{<{}<}, \texttt{**}, \texttt{/}, \texttt{\%}, \texttt{>{}>}, \texttt{\&}, \texttt{|}, \texttt{$^\wedge$}, \texttt{==}, \texttt{!=}, \texttt{<}, \texttt{<=}, \texttt{>}, and \texttt{>=}. 

Full documentation of these operators can be found in the Tcl \hyperlink{https://www.tcl-lang.org/man/tcl/TclCmd/mathop.htm}{::tcl::mathop namespace documentation}.

\begin{syntax}
\command{.\$op} \$ndlist ...
\end{syntax}
\begin{args}
\$op & Tcl numeric math operator. \\
\$ndlist ... & Numeric ND-lists to apply operator to.
\end{args}

\begin{example}{Prefix notation math}
\begin{lstlisting}
namespace path {::tcl::mathfunc ::tcl::mathop}; # exposes 'expr' functions and operators
# Scalar math
set x [+ 1 2]; # 3
puts [** $x 2]
# Element-wise math
set x [.+ {10 20} [nreshape [range 6] {2 3}]]; # {10 11 12} {23 24 25}
puts [napply double [.- [nget $x 0* 0:end-1]]]
\end{lstlisting}
\tcblower
\begin{lstlisting}
9
-10.0 -11.0
\end{lstlisting}
\end{example}

\clearpage